Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-09-20 documents/mixture_models.tex
clarification, simplification

The value of these mixture-of-Gaussian galaxy components comes when
they are to be convolved with a PSF (usually in fact a pixel-convolved
PSF) that is itself represented as a mixture of Gaussians.  In this
scenario, the PSF $\psi(\Delta\pos)$---which is thought of as a
function of focal-plane displacement $\Delta\pos$ away from, say, a
true stellar position---is represented as a $K$-Gaussian mixture
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: It implies that a star of flux $S_s$ at focal-plane
position $\pos_s$ will lead to an image (PSF-convolved intensity map)
of the form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a mixture of Gaussians, any
image of a star---or indeed any image of any set of stars---is also
represented as a mixture of Gaussians.

Applying this PSF to an exponential or de~Vaucouleurs galaxy is
slightly more complicated, because the galaxy has not just a flux
$S_g$ and a central position $\pos_g$; it also has a shape.  Because
we are only considering these simple galaxies, we are only permitting
ellipsoidal shapes, which can be represented by a semi-major axis $a$,
a semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or by a symmetric
variance tensor $\var_g$ (which is a general representation of an
ellipse).  The galaxy is distorted by this variance tensor
\emph{prior} to PSF convolution, so the focal-plane image
(PSF-convolved intensity field) for a general exponential galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\var_g) &=& \sum_{k=1}^K \sum_{m=1}^{M_g} S_g\,g_m\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var_m\cdot\transpose{\affine_g}
\\
\var_g &\equiv& \affine_g\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Since implicitly all vectors are
two-dimensional column vectors, this makes $\affine_g$ a $2\times 2$
affine transformation matrix and $\var_g$ a symmetric $2\times 2$
variance tensor.  Similarly, the general de~Vaucouleurs galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\dev,S_h,\pos_h,\var_h) &=& \sum_{k=1}^K \sum_{m=1}^{M_h} S_h\,h_m\,p_k\,\normal(\pos\given\pos_h+\mean_k,\var_{hn}+\var_k)
\\
\var_{hn} &\equiv& \affine_h\cdot\var_m\cdot\transpose{\affine_h}
\\
\var_h &\equiv& \affine_h\cdot\transpose{\affine_h}
\quad ,
\end{eqnarray}
where the matrix mathematics is the same as that in the exponential
case.  Note the important and key result of this \documentname, to
wit, that a mixture-of-Gaussian galaxy model (with $M$ components)
convolved with a mixture-of-Gaussian PSF model (with $K$ components)
yields a mixture-of-Gaussian image (with $[M\,K]$ components).

>>>>

The value of these mixture-of-Gaussian galaxy components comes when
they are to be convolved with a PSF (usually in fact a pixel-convolved
PSF) that is itself represented as a mixture of Gaussians.  In this
scenario, the PSF $\psi(\Delta\pos)$---which is thought of as a
function of focal-plane displacement $\Delta\pos$ away from, say, a
true stellar position---is represented as a $K$-Gaussian mixture
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: A star of flux $S_s$ at focal-plane position
$\pos_s$ will lead to an image (PSF-convolved intensity map) of the
form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a mixture of Gaussians, any
image of a star---or indeed any image of any set of stars---is also
represented as a mixture of Gaussians.

Applying this PSF to an exponential or de~Vaucouleurs galaxy is
slightly more complicated, because the galaxy has not just a flux
$S_g$ and a central position $\pos_g$; it also has a shape.  Because
we are only considering these simple galaxies, we are only permitting
ellipsoidal shapes, which can be represented by a semi-major axis $a$,
a semi-minor axis $b$, and a position angle $\phi$, or equivalently by
eigenvalues $a, b$ and eigenvectors $\uv_1, \uv_2$, or by a symmetric
variance tensor $\var_g$ (which is a general representation of an
ellipse).  The galaxy is distorted by this variance tensor
\emph{prior} to PSF convolution, so the focal-plane image
(PSF-convolved intensity field) for a general exponential galaxy is
given by
\begin{eqnarray}\displaystyle
I(\pos\given\exp,S_g,\pos_g,\affine_g) &=& \sum_{k=1}^K \sum_{m=1}^{M_{\exp}} S_g\,a_{\exp,m}\,p_k\,\normal(\pos\given\pos_g+\mean_k,\var_{gm}+\var_k)
\\
\var_{gm} &\equiv& \affine_g\cdot\var_{\exp,m}\cdot\transpose{\affine_g}
\\
\affine_g &=& \left[a\,\uv_1 , b\,\uv_2 \right]
\quad ,
\end{eqnarray}
where $a$ and $b$ are the major and minor axis lengths of the galaxy
ellipse (in appropriate units) and $\uv_1$ and $\uv_2$ are the
eigenvectors on the sky pointing in the major-axis and minor-axis
directions respectively.  Implicitly all vectors are two-dimensional
column vectors, and $\affine_g$ is a $2\times 2$ affine transformation
matrix that contains the ``shape'' (position angle, major-axis, and
ellipticity) information about the galaxy.  The $\dev$, $\luv$, and
$\lux$ cases are all similar.  Note the important and key result of
this \documentname, to wit, that a mixture-of-Gaussian galaxy model
(with $M$ components) convolved with a mixture-of-Gaussian PSF model
(with $K$ components) yields a mixture-of-Gaussian image (with
$[M\,K]$ components).
